---
title: "Classification and Supervised Machine Learning"
output: pdf_document
editor_options: 
  chunk_output_type: inline
---
Using Machine Learning to Predict Subscription to Bank Term Deposits for Clients 

Marketing to potential clients has always been a crucial challenge in attaining success for banking institutions. It’s not a surprise that banks usually deploy mediums such as social media, customer service, digital media and strategic partnerships to reach out to customers. But how can banks market to a specific location, demographic, and society with increased accuracy? With the inception of machine learning - reaching out to specific groups of people have been revolutionized by using data and analytics to provide detailed strategies to inform banks which customers are more likely to subscribe to a financial product. In this project on bank marketing with machine learning, I will explain how a particular Portuguese bank can use predictive analytics from data science to help prioritize customers which would subscribe to a bank deposit.

The data set is based off the direct marketing campaigns of a Portuguese banking institution. These marketing campaigns were based on phone calls. More than one contact to a client was required, in order to know if the product (bank term deposit) was subscribed by a client or not. The classification goal is to predict if a client will subscribe to the bank term deposit (yes/no).

The dataset contains 21 columns including the output (y). I am going to discard the output column and use the remaining columns to find the most relatable independent variables (x) that will be able to predict if a customer will subscribe to a bank deposit or not.

In this project I will demonstrate how to build a model predicting clients subscribing to a bank's term deposit in R using the following steps:

-data exploration

-feature engineering

-building training/validation/test samples

-model selection

-model evaluation
```{r}
library(readr)
library(dplyr)
library(tidyverse)
#install.packages("caret")
library(caret)
#install.packages("pROC")
library(pROC)
#install.packages("rpart")
library(rpart)
#install.packages("rpart.plot")
library(rpart.plot)
#install.packages("PRROC")
library(PRROC)
#install.packages('randomForest')
library(randomForest)
```
Partitioning the data into training and testing datasets.
```{r}
marketing <- read_csv("https://raw.githubusercontent.com/domrongphol/SPE487/master/Homework/HW2/bank-full.csv")
set.seed(123)
random_sample <- createDataPartition(marketing$y,p = 0.8, list = FALSE)
# Generating training dataset  
training_dataset  <- marketing[random_sample, ] 

# Generating testing dataset
testing_dataset <- marketing[-random_sample, ] 
training_dataset$y <- factor(training_dataset$y, levels = c("no", "yes"))
testing_dataset$y <- factor(testing_dataset$y, levels = c("no", "yes"))


# downsample imbalanced data set
train_balanced <- downSample(x = training_dataset[, -ncol(training_dataset)], 
                             y = training_dataset$y)
names(train_balanced)[ncol(train_balanced)] <- "y"
table(train_balanced$y)

logit_model <- glm(y ~ age + job + marital + education + default + balance + housing + loan + duration +
campaign + pdays + previous + poutcome, data = train_balanced, family = binomial)

# predicting the target variable
train_predictions <- predict(logit_model, train_balanced, type = "response")
predictions <- predict(logit_model, testing_dataset, type = "response")

# Convert probabilities to class labels based on a 0.5 threshold
predicted_classes <- ifelse(predictions > 0.5, "yes", "no")
predicted_classes_l <- ifelse(train_predictions > 0.5, "yes", "no")

# Confusion Matrix
conf_matrix_train <- confusionMatrix(as.factor(predicted_classes_l), as.factor(train_balanced$y))
conf_matrix <- confusionMatrix(as.factor(predicted_classes), as.factor(testing_dataset$y))

# Generate the ROC curve object
roc_curve_train <- roc(train_balanced$y, train_predictions)
roc_curve <- roc(testing_dataset$y, predictions)

# Plot the ROC curve
roc_logit_train <- plot(roc_curve_train, main = "ROC training data", col = "blue", lwd = 2)
roc_logit <- plot(roc_curve, main = "ROC Curve for Logistic Regression", col = "blue", lwd = 2)

# Calculate AUC
auc_value_train <- auc(roc_logit_train)
auc_value <- auc(roc_curve)
```
The data partition choice is 80-20 to ensure there is enough data for model training while keeping enough for robust testing. 

The logit model has a reasonably high performance overall, with  accuracy of 83.45%, this makes sense as the test data set is imbalanced as it can be dominated by the majority class 'no'. For the 'no' class, sensitivity indicates how well the model identifies true 'no' instances. Here, it captures 84.57% of the actual 'no' cases.For the 'yes' class, specificity is lower, indicating that the model correctly identifies only 75.02% of 'yes' instances.

The AUC of 0.8728 demonstrates good discriminatory ability between the two classes, indicating that the model generally performs well in distinguishing between "yes" and "no" outcomes.


```{r}
# Train the decision tree 
dec_tree_model <- rpart(y ~ age + job + marital + education + default + balance + housing + loan + duration + campaign + pdays + previous + poutcome, 
                    data = train_balanced, 
                    method = "class")

# Visualize the decision tree
rpart.plot(dec_tree_model, main = "Decision Tree for Subscription Classification")

# Model Evaluation
# Ensure the predicted values are factors with the same levels as the actual values
train_pred_prob <- predict(dec_tree_model, train_balanced, type = "prob")[,2]
train_pred_tree <- ifelse(train_pred_prob > 0.3, "yes", "no")
train_pred_tree <- factor(train_pred_tree, levels = levels(train_balanced$y))
confusionMatrix(train_pred_tree, train_balanced$y)

# For testing data, making sure the factor levels match as well
test_pred_prob <- predict(dec_tree_model, testing_dataset, type = "prob")[,2]
  test_pred_tree <- ifelse(test_pred_prob > 0.3, "yes", "no")
test_pred_tree <- factor(test_pred_tree, levels = levels(testing_dataset$y))
confusionMatrix(test_pred_tree, testing_dataset$y)


# ROC curve 
train_roc_dt <- roc(train_balanced$y, train_pred_prob)
test_roc_dt <- roc(testing_dataset$y, test_pred_prob)

roc_dt <- plot(roc_curve, main = "ROC Curve for Decision Tree Model", col = "blue", lwd = 2)

# AUC
test_auc_value <- auc(test_roc_dt)
train_auc_value <- auc(train_roc_dt)
```
By lowering the threshold, the model is more likely to classify cases as "yes". This means that the model will flag more potential positives (the "yes"), which increases recall. By reducing the threshold to 0.3, the model is optimizing for recall at the cost of precision, which may be appropriate as we are trying to ensure that no potential positive cases are missed, even if it leads to more false alarms.

AUC of 0.81 (close to 1) indicates a fairly well performing model.

```{r}
printcp(dec_tree_model)
best_cp <- dec_tree_model$cptable[which.min(dec_tree_model$cptable[,"xerror"]), "CP"]
pruned_model <- prune(dec_tree_model,cp = best_cp)

rpart.plot(pruned_model, main = "Pruned Decision Tree for Subscription Classification")
# apply pruned tree to test data
test_pred_pruned <- predict(pruned_model, testing_dataset, type = "class")

# Ensure the predicted values have the same factor levels as the actual test labels
test_pred_pruned <- factor(test_pred_pruned, levels = levels(testing_dataset$y))

# Confusion Matrix for Pruned Model on Test Data
confusionMatrix(test_pred_pruned, testing_dataset$y)
```

Pruning selects the lowest complexity parameter value associated with a shorter tree that minimizes the cross-validated error. The pruned tree contains 5 splits. It looks like the pruned and unpruned model have mostly identical results, indicating that the unpruned model is well-tuned.


```{r}
forest <- randomForest(y ~ age + job + marital + education + default + balance + housing + loan + duration + campaign + pdays + previous + poutcome, 
                    data = train_balanced, ntree=2000)
# Variable importance
varImpPlot(forest)

# Predict on training and testing datasets
train_pred_rf <- predict(forest, train_balanced)
test_pred_rf <- predict(forest, testing_dataset, type = "prob")

# Apply the 0.3 threshold
test_pred_rf_prob <- predict(forest, testing_dataset, type = "prob") 
test_pred_rf <- ifelse(test_pred_rf_prob[,2] >= 0.3, "yes", "no")
test_pred_rf <- factor(test_pred_rf, levels = levels(testing_dataset$y))
# Confusion Matrix for training and testing data
train_conf_matrix <- confusionMatrix(train_pred_rf, train_balanced$y)
test_conf_matrix <- confusionMatrix(test_pred_rf, testing_dataset$y)

# ROC and AUC 
train_pred_rf_prob <- predict(forest, train_balanced, type = "prob")
train_roc_curve <- roc(train_balanced$y, train_pred_rf_prob[,2])
train_auc_value <- auc(train_roc_curve)
test_pred_rf_prob <- predict(forest, testing_dataset, type = "prob")
test_roc_rf <- roc(testing_dataset$y, test_pred_rf_prob[,2])
test_auc_rf <- auc(roc_curve)

# Plot ROC curve
roc_rf <- plot(roc_curve, main = "ROC Curve for Random Forest Model", col = "blue", lwd = 2)
```
It looks like the variable "duration" is the most important variable in predicting our outcome variable. The next most influential variables are "balance" and "age" but they are far behind the "duration" variable based on the plot.

The training model correctly predicted 94.8% of the true negatives, and 95.25% of the true positives, with high (95%) precision rate as well as high (94.82%) negative predictive value. The testing model correctly predicted 70.1% of the true negatives, and 90.73% of the true positives. It has a very high (98.28%) precision rate but low (28.73%) negative predictive value. The AUC is 0.8859, implying good discrimination and predictability.

A low threshold was chosen to increase the number of true positives in the model.

```{r}
# Define the tuning grid with possible values for mtry
tune_grid <- expand.grid(mtry = c(2, 3, 4, 5, 6)) 

# Train the Random Forest model with tuning using 10 k fold
train_control <- trainControl(method = "cv", number = 5)

rf_model <- train(y ~ age + job + marital + education + default + balance + housing + loan + duration + campaign + pdays + previous + poutcome,
                  data = train_balanced,
                  method = "rf",
                  trControl = train_control,
                  tuneGrid = tune_grid)

# model using mtry = 5
# Train the random forest model with mtry = 5
forest_2 <- randomForest(y ~ age + job + marital + education + default + balance + housing + loan + duration + campaign + pdays + previous + poutcome, 
                         data = train_balanced, mtry = 5)

# Make predictions on the training set
train_pred_rf_2 <- predict(forest_2, train_balanced)

# Make predictions on the testing set (probabilities)
test_pred_rf_2 <- predict(forest_2, testing_dataset, type = "prob")

# Apply the 0.3 threshold to the predicted probabilities
test_pred_rf_2_class <- ifelse(test_pred_rf_2[,2] >= 0.3, "yes", "no")
test_pred_rf_2_class <- factor(test_pred_rf_2_class, levels = levels(testing_dataset$y))

# Confusion Matrix for training and testing data
train_conf_matrix_2 <- confusionMatrix(train_pred_rf_2, train_balanced$y)
test_conf_matrix_2 <- confusionMatrix(test_pred_rf_2_class, testing_dataset$y)

# Compute ROC and AUC 
roc_curve <- roc(testing_dataset$y, test_pred_rf_2[, 2])  # Use probabilities from the model forest_2
auc_value <- auc(roc_curve)

# Plot ROC curve
plot(roc_curve, main = "ROC Curve for Random Forest Model (mtry = 5)", col = "blue", lwd = 2)

# Output AUC value
auc_value
```
I chose to tune the mtry parameter because it controls the number of features that are considered when splitting each node in the decision trees. A larger value of mtry, as opposed to the original value 3, reduces bias, but increases variance. By considering more variables it captures different complex interactions between variables that might affect deposit subscriptions.

8. Use a Categorical Boosting (CATBoost) model to classify your data using the original formula. Classify
your training data and testing data. After that, you need to create a confusion matrix, and calculate
accuracy, precision, and recall. Run the ROC graph and calculate AUC. Interpret your ROC graph
and the calculated AUC.

```{r}
#install.packages('remotes')
#remotes::install_url(
  'https://github.com/catboost/catboost/releases/download/v1.2.7/catboost-R-darwin-universal2-1.2.7.tgz',
  INSTALL_opts = c("--no-multiarch", "--no-test-load", "--no-staged-install")
)

#install.packages("devtools")
library(catboost)

training_dataset[] <- lapply(training_dataset, function(x) {
  if (is.character(x)) as.factor(x) else x
})

testing_dataset[] <- lapply(testing_dataset, function(x) {
  if (is.character(x)) as.factor(x) else x
})


# Convert the target variable to numeric (1 for "yes", 0 for "no")
training_dataset$y <- ifelse(training_dataset$y == "yes", 1, 0)
testing_dataset$y <- ifelse(testing_dataset$y == "yes", 1, 0)
train_labels <- training_dataset$y
test_labels <- testing_dataset$y

# convert data to matrices
train_data <- data.matrix(training_dataset[, -which(names(training_dataset) == "y")])
test_data <- data.matrix(testing_dataset[, -which(names(testing_dataset) == "y")])


# Identify categorical features and adjust for zero-indexing
cat_features <- which(sapply(training_dataset[, -which(names(training_dataset) == "y")], is.factor)) - 1

# restart R environment
#rm(list = ls())

# Create CatBoost pools with categorical features specified
train_pool <- catboost.load_pool(data = train_data, label = train_labels, cat_features = cat_features)
test_pool <- catboost.load_pool(data = test_data, label = test_labels, cat_features = cat_features)

# Train the CatBoost model
params <- list(loss_function = "Logloss", iterations = 100, depth = 6, learning_rate = 0.1, verbose = 1)
model <- catboost.train(train_pool, params = params)

# Predictions
train_pred <- catboost.predict(model, train_pool, prediction_type = "Probability")
test_pred <- catboost.predict(model, test_pool, prediction_type = "Probability")

# Evaluate accuracy
train_accuracy <- mean(train_pred == train_labels)
test_accuracy <- mean(test_pred == test_labels)

roc(test_labels, test_pred)
roc_cat <- plot(roc_curve, main = "ROC for CatBoost", col = "blue", lwd =2)

auc(roc_curve)
```
Model Evaluation

Given that the model is imbalanced, accuracy may not be the best metric. A combination of the precision and recall metric, the F-1 score, would be an ideal metric for comparison as it balances both false positives and false negatives. AUC-ROC also provides a good measure to compare model performance across different probability thresholds for the dataset.

```{r}
model_comparison <- data.frame(
 Model = c("Logistic Regression", "Decision Tree", "Random Forest", "CATBoost"),
   AUC_ROC = c(0.8728,0.81,0.8859, 0.8809))
model_comparison
```
The random forest appears to be the top performer on testing data.
